---
title: "Functional Enrichment Analysis"
author: "Duanchen Sun, Ling-Yun Wu"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Functional Enrichment Analysis}
  %\VignetteEngine{knitr::rmarkdown}
---


# NetGen: Network-based Generative Model

NetGen is a network-based generative model for functional enrichment analysis. We first load the required packages

```{r}
library(CopTea)
options(scipen=0)
```

In this example, we use a small subset of GO annotation database, which contains 300 biological processes categories.
```{r}
load("GO_BP_300.RData")
```
The annotation data is stored in a matrix with dimensions:
```{r}
dim(annotation)
```
which indicates there are total 6447 genes and 300 GO terms in this annotation dataset.

Next, we load the protein-protein interaction (PPI) dataset:
```{r}
load("PPI.RData")	
```
The PPI network is given by its adjacent matrix as follows:
```{r}
dim(adj_matrix)
```

We simulate the gene list of interest:
```{r}
load("active_gene.RData")
```
The list consists of 84 active genes which are derived from the true categories as follows.
```{r}
True_Categories <- c("GO:0019614", "GO:1903249", "GO:2000506", "GO:0015985",
                     "GO:0071962")
```

## Mode1: fixed parameter strategy

In the first part of this example, we try to identify the most enriched categories using a given parameter setting.
```{r}
Enriched_Categories <- netgen(annotation, adj_matrix, active_gene,
                              p1 = 0.8, p2 = 0.1, q = 0.001, alpha = 5, trace=TRUE)
```

The most enriched categories identified by NetGen are
```{r}
Enriched_Categories
```
and the false negative categories are
```{r}
setdiff(True_Categories, Enriched_Categories[,1])
```

## Mode2: mixed parameter strategy

Instead of using a fixed parameter setting, we can run NetGen with several different parameter settings,
and then select the result of highest enrichment significance.
```{r}
p1 <- c(0.5, 0.8)
p2 <- c(0.1, 0.3)
q  <- 0.001
Enriched_Categories <- netgen(annotation, adj_matrix, active_gene,
                              p1, p2, q, alpha = 3, trace=FALSE)
```
The combined p-values of mixed parameter strategy are
```{r}
Enriched_Categories$Term_combined_pvalue
```
And the most enriched categories and its corresponding parameter combination are
```{r}
Enriched_Categories$mix_result[which.min(Enriched_Categories$Term_combined_pvalue)]
```
